Reflections on reflection for \\ intuitionistic set theories

### 2019-01-10 paper-reflection.tex
Proof of the reflection principle for IZF+RRS

\newcommand{\ZF}{\textsc{zf}}
\newcommand{\IZF}{\textsc{izf}}
\newcommand{\CZF}{\textsc{czf}}
\newcommand{\ZFC}{\textsc{zfc}}
\newcommand{\ZFCS}{\textsc{zfc/s}}
\newcommand{\RRS}{\textsc{rrs}}
\newcommand{\ES}{(\IZF+\RRS)/\textsc{s}}

@@

>>>>

\newcommand{\ZF}{\textsc{zf}}
\newcommand{\IZF}{\textsc{izf}}
\newcommand{\CZF}{\textsc{czf}}
\newcommand{\ZFC}{\textsc{zfc}}
\newcommand{\ZFCS}{\textsc{zfc/s}}
\newcommand{\RRS}{\textsc{rrs}}
\newcommand{\MDC}{\textsc{mdc}}
\newcommand{\ES}{(\IZF+\RRS)/\textsc{s}}

@@

\section{Constructivizing the reflection principle}
\label{sect:constructive-proof}

Our constructive rendition of the reflection principle will require Aczel's
relation reflection scheme~$\RRS$ displayed in Definition~\ref{defn:rrs}. This
result is the best possible, as conversely the reflection principle entails the
relation reflection scheme. The relation reflection scheme first surfaced in
the theory of coinductive definitions of classes and enjoys substantial
stability properties, as it passes from the metatheory to XXX[all kinds of]
models. Background on~$\RRS$ can be found in Aczel's original article
introducing it~\cite{aczel:rrs}.

\begin{defn}[$\RRS$, Aczel's relation reflection scheme]\label{defn:rrs}
Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$A \subseteq X$ be a
subset. If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
then there is a set~$B$ such that~$A \subseteq B \subseteq X$ and such
that~$\forall x \in B\_ \exists y \in B\_ \langle x,y \rangle \in R$.
\end{defn}

For our purposes, Palmgren's multivalued dependent choice is slightly more
convenient to work with than~$\RRS$. It is equivalent, over~$\CZF$ and
a~fortiori over~$\IZF$, to~$\RRS$~\cite{palmgren:mdc}.

\begin{defn}[$\MDC$, Palmgren's multivalued dependent choice]
Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$A \subseteq X$ be a
subset. If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
then there is a function~$f : \NN \to P(X)$ (the class of all subsets of~$X$)
such that~$A \subseteq f(0)$ and such that~$\forall x \in f(n)\_ \exists y \in
f(n+1)\_ \langle x,y \rangle \in R$ for every~$n$.\end{defn}

Even though superficially similar, the following lemma is \emph{not} yet a
constructivization of Lemma~\ref{lemma:zf-smallstep}; these two lemmas differ
in the set from which~$u_1,\ldots,u_n$ are drawn.

\begin{lemma}\label{lemma:izf-microstep}
Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
theory with free variables as indicated. Then~$\IZF$ proves
\begin{multline*}
  {\qquad}\forall H\_
  \exists H' \supseteq H\_
  \forall u_1,\ldots,u_n \in H\_ \\
  \begin{array}{@{}rcl@{}}
  (\exists x\_ \varphi(u_1,\ldots,u_n,x)) &\Longrightarrow&
  (\exists x \in H'\_ \varphi(u_1,\ldots,u_n,x)) \quad\mathop{\wedge} \\
  (\forall x\_ \varphi(u_1,\ldots,u_n,x)) &\Longleftarrow&
  (\forall x \in H'\_ \varphi(u_1,\ldots,u_n,x)).\end{array}
  {\qquad}
\end{multline*}
Furthermore, we may assume that~$H'$ is transitive and closed under subsets;
and given a finite list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of
the single formula~$\varphi$, we may suppose that~$H'$ has the displayed
property for each of the formulas~$\varphi_i$.
\end{lemma}

\begin{proof}Let~$\Omega \defeq P(\{0\})$ be the set of truth values.
For given elements~$u_1,\ldots,u_n \in H$, we have
\[ \begin{array}{@{}l@{}l@{}}
  \forall a \in \{ a \in \{0\} \,|\, \exists x\_ \varphi(u_1,\ldots,u_n,x) \}\_ &
  \exists x\_ \varphi(u_1,\ldots,u_n,x) \quad\text{and} \\
  \forall p \in \{ p \in \Omega \,|\, \exists x\_ (0 \in p \Leftrightarrow
  \varphi(u_1,\ldots,u_n,x)) \}\_ &
  \exists x\_ (0 \in p \Leftrightarrow \varphi(u_1,\ldots,u_n,x)).
\end{array} \]
Hence, by collection, there are sets~$C$ and~$D$ such that
\[ \begin{array}{@{}rcl@{}}
  (\exists x\_ \varphi(u_1,\ldots,u_n,x)) &\Longrightarrow&
  (\exists x \in C\_ \varphi(u_1,\ldots,u_n,x)) \quad\text{and} \\
  (\forall x\_ \varphi(u_1,\ldots,u_n,x)) &\Longleftarrow&
  (\forall x \in D\_ \varphi(u_1,\ldots,u_n,x)).
\end{array} \]
The union~$C \cup D$ satisfies both of these conditions at once.

Applying collection again, there is a set~$X$ such that for
any~$u_1,\ldots,u_n \in H$ there exists a set~$E \in X$ such that
\[ \begin{array}{@{}rcl@{}}
  (\exists x\_ \varphi(u_1,\ldots,u_n,x)) &\Longrightarrow&
  (\exists x \in E\_ \varphi(u_1,\ldots,u_n,x)) \quad\text{and} \\
  (\forall x\_ \varphi(u_1,\ldots,u_n,x)) &\Longleftarrow&
  (\forall x \in E\_ \varphi(u_1,\ldots,u_n,x)).
\end{array} \]
Hence the set~$H' \defeq M \cup \bigcup X$ has the required property.

To ensure that~$H'$ is transitive and closed under subsets, we simply pass
first to its transitive closure and then compute the union of all its
finitely-iterated powersets.

In order to accomodate more than a single formula~$\varphi$, we add one summand
in the definition of~$H'$ for each formula~$\varphi_i$.
\end{proof}

\begin{lemma}\label{lemma:izf-smallstep}
Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
theory with free variables as indicated. Then~$\IZF+\RRS$ proves
\begin{multline*}
  {\qquad}\forall M\_
  \exists S \supseteq M\_
  \forall u_1,\ldots,u_n \in S\_ \\
  \begin{array}{@{}rcl@{}}
  (\exists x\_ \varphi(u_1,\ldots,u_n,x)) &\Longrightarrow&
  (\exists x \in S\_ \varphi(u_1,\ldots,u_n,x)) \quad\mathop{\wedge} \\
  (\forall x\_ \varphi(u_1,\ldots,u_n,x)) &\Longleftarrow&
  (\forall x \in S\_ \varphi(u_1,\ldots,u_n,x)).\end{array}
  {\qquad}
\end{multline*}
Furthermore:
\begin{enumerate}
\item We may suppose that~$S$ is transitive.
\item We may suppose that~$S$ is closed under subsets.
\item Given a finite list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of
the single formula~$\varphi$, we may suppose that~$S$ bounds all of
the formulas~$\varphi_i$.
% XXX: better explanation
\end{enumerate}
\end{lemma}

\begin{proof}By~$\MDC$ and Lemma~\ref{lemma:izf-microstep}, there is a
function~$f : \NN \to P(V)$, where~$V$ is the universe, such that~$M \in
f(0)$ and such that for any number~$k$ and any set~$H \in f(k)$, there is a
set~$H' \in f(k+1)$ such that~$H$ and~$H'$ are related as in the conclusion of
Lemma~\ref{lemma:izf-microstep}.

We set~$S \defeq \bigcup \bigl(\bigcup_{k \in \NN} f(k)\bigr)$. This set~$S$
has the required property; to verify this, it is useful to observe that
given~$u_1,\ldots,u_n \in S$, there exists a common number~$k$ such
that~$u_1,\ldots,u_n \in H$ for some~$H \in f(k)$.

In order to ensure addendum~(1) and~(2), we apply~$\MDC$ in a slightly
different way such that for any number~$k$ and any set~$H \in f(k)$, there is a
set~$H' \in f(k+1)$ such that~$H$ and~$H'$ are related as in the conclusion of
Lemma~\ref{lemma:izf-microstep} and such that~$H'$ is transitive and closed
under subsets. Even though it cannot be expected that for any number~$k$,
any set~$H \in f(k)$ will be transitive and closed under subsets, the union~$S$
will.
\end{proof}

\begin{thm}\label{thm:izf-bigstep}
Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory with free variables as indicated. Then~$\IZF+\RRS$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
\end{thm}

\begin{proof}The proof of Theorem~\ref{thm:zf-bigstep} carries over. The only
difference is that instead of Lemma~\ref{lemma:zf-smallstep},
Lemma~\ref{lemma:izf-smallstep} has to be used, and that the case for the
universal quantifier has to be treated just as the case for the existential
quantifier is.\end{proof}
